---
title: "Rules for field/column combinations"
author: "Edwin van Leeuwen, Marc Baguelin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r,include=F}
library(pander)
```

The package `cleanr` provide tools to efficiently check the integrity and consistency of datasets. In particular, set of rules to which fields/columns need to conform to can be implemented. Rules can be as simple as checking that the values in one column are always be higher than in another or more complex functional between values of two fields. The defined rules can be read from a yaml file, or passed as a list.

There are two types of rules supported 1) a required relation between fields (e.g. one field need to always be higher than the other and 2) accepted/valid and invalid combinations of values (e.g. the outcome of a disease needs to be death if a time of death is set).

Below follow some examples. All examples use an simulated set of epidemiological data (`epi.sim`)
```{r,echo=F}
library(cleanr)
data(epi.sim)
pander(head(epi.sim))
```

## Relation between columns

Relation rules are linking two columns with a rule that should be respected for all the pairs of values. To test if the rule is respected in a given dataset, the `rule_validation` function can be used with the rules inputed from a yaml file.

For example by using the rule.yaml file that contains the following rule:
```
- fields:
    - date.of.onset
    - date.of.hospitalisation
  relation: lte
```
we require that date.of.onset should always be lesser than or equal to date.of.hospitalisation.
```{r,eval=F}
rule_validation(epi.sim, rules_file="rule.yaml")
```
which passes for our data set. In contrast, if we require the date.of.onset to be strictly lesser than, by replacing `lte` with `lt` it will throw a warning and return `FALSE`:
```{r,echo=F}
  data(epi.sim)
  rule_validation(epi.sim, rules=list(list(
    "fields"=c("date.of.onset", "date.of.hospitalisation"), "relation"="lt" ) ) )
```

## (In)valid combinations of values

[in progress]

The combination set of rules define combinations of values between columns which are valid or invalid.  
For example by using the rule.yaml file that contains the following rule:
```
- fields:
    - gender
    - generation
  invalid: [f,0]
```

```{r,echo=F}
  data(epi.sim)
  rule_validation(epi.sim, rules=list(list(
    "fields"=c("gender", "generation"), "invalid"=c("f",0) ) ) )
```

Similarly if we specify `invalid: [f,100]` it does pass, because this combination does not occur in the data set.

You can also provide all the valid combinations by using `valid: [["f",1],["m",1], ...]`. Be aware that in this case you need to specify all the valid combinations and it will throw an error if the data contains any information that is not there.

## Conditional values in dataset

Conditional is more general version of valid/invalid although limited to two fields.
Some values are imposed conditional of others in the dataset. For example if a patient is given a date of death, the outcome  should be "death". A patient who has a different status than `death` while having a date of death should be flagged up for correction.

A rule testing for that would be: 
```
- fields: [date.of.death, outcome]
  conditional:
    operator: ["not", "equal"]
    values: ["NA", "Death"]
```
i.e. if the date.of.death is not equal to NA the outcome has to be equal to Death.
 
In the `epi.sim` dataset quite a few records have a data of death recorded, so testing it for this rule results in a warning.

```{r,eval=F}
rule_validation(epi.sim, rules_file = rules.yaml )
```

```{r,echo=F}
  rules <- list(list("fields"=c("date.of.death","outcome"),
              "conditional"=list("operator" = c("not","equal"), "values"=c("NA","Death"))))
 rule_validation(epi.sim, rules = rules )
```

## Operators

For both the relation and conditional rules a number of operators are supported:

  - equal, ==
  - not, !=
  - gt, >, greater_than
  - lt, <, lesser_than
  - lte, <=, lesser_than_or_equal
  - gte, >=, greater_than_or_equal
 
